---
title: "Hurray! My webpage is finished!"
author: "Jakub Kuzilek"
date: 2018-01-23
categories: ["personal"]
tags: ["academic"]
summary: "Finally, after days of struggling with bureaucracy, I managed to find enough time to complete my first `blogdown` project - my personal academic webpage! I must admit it `blogdown` is awesome and the way Hugo works is brilliant. "
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE)
```

Finally, after days of struggling with bureaucracy, I managed to find enough time to complete my first `blogdown` project - my personal academic webpage! I must admit it `blogdown` is awesome and the way Hugo works is brilliant. 

What seems to be problematic is to migrate all my publications list. I think I will start with that. Maybe this will be my next post in near future (hope not far future...).

Btw. I really like the pyramid bar chart, which was included in a template file. Thus I will keep it here:

```{r pie, fig.cap='A fancy pie chart.', tidy=FALSE}
par(mar = c(0, 1, 0, 1))
pie(
  c(280, 60, 20),
  c('Sky', 'Sunny side of pyramid', 'Shady side of pyramid'),
  col = c('#0292D8', '#F7EA39', '#C4B632'),
  init.angle = -50, border = NA
)
```
